---
title: "Evaluating Branched Pseudotime Trajectories with BioTrajX"
output: github_document
vignette: >
  %\VignetteIndexEntry{Evaluating Branched Pseudotime Trajectories with BioTrajX}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  warning = FALSE,
  message = FALSE,
  eval = FALSE   # requires downloaded data — run interactively after loading stem_cell.rds
)
```

This article walks through applying BioTrajX's DOE metrics to a branched
haematopoietic stem cell differentiation dataset: aging mouse hematopoietic
stem and progenitor cells, where the `Stem_Progenitors` compartment bifurcates
into an erythroid lineage (`Erythroid_progenitors_Erythroblasts` →
`Erythrocytes`) and a B-cell lineage (`Immature_B` → `Mature_B`). As in the
linear-trajectory vignette, pseudotime from eight TI methods is already
computed and stored as `meta.data` columns.

## 1. Load the data

```{r load}
library(BioTrajX)
library(Seurat)

stem <- readRDS("data/stem_cell.rds")
stem
```

`stem$Phenotype` holds the ten cell-type labels, and the same eight
TI-method pseudotime columns used in the linear-trajectory vignette
(`Slingshot`, `CytoTRACE`, `Monocle3`, `DPT`, `SCORPIUS`, `TSCAN`,
`PAGA-DPT`, `Palantir`) are already present in `meta.data`.

## 2. Visualize branch labels and pseudotime

One UMAP panel for the `Phenotype` labels, plus one panel per TI method colored by that method's pseudotime:

```{r visualize}
pt_cols <- c("Slingshot", "CytoTRACE", "Monocle3", "DPT",
             "SCORPIUS", "TSCAN", "PAGA-DPT", "Palantir")

umap_df <- as.data.frame(Embeddings(stem, "umap"))
colnames(umap_df) <- c("UMAP1", "UMAP2")
umap_df$CellType <- stem$Phenotype[rownames(umap_df)]

p_celltype <- ggplot2::ggplot(umap_df, ggplot2::aes(UMAP1, UMAP2, colour = CellType)) +
  ggplot2::geom_point(size = 0.5, alpha = 0.7) +
  ggplot2::theme_classic(base_size = 10) +
  ggplot2::labs(title = "Cell type", colour = NULL)

method_plots <- lapply(pt_cols, function(method) {
  df <- cbind(umap_df[, c("UMAP1", "UMAP2")], Pseudotime = stem@meta.data[[method]])
  ggplot2::ggplot(df, ggplot2::aes(UMAP1, UMAP2, colour = Pseudotime)) +
    ggplot2::geom_point(size = 0.5, alpha = 0.7) +
    ggplot2::scale_colour_distiller(palette = "Spectral", direction = -1,
                                     na.value = "grey80") +
    ggplot2::theme_classic(base_size = 10) +
    ggplot2::labs(title = method)
})

if (requireNamespace("patchwork", quietly = TRUE)) {
  patchwork::wrap_plots(c(list(p_celltype), method_plots), ncol = 3)
} else {
  p_celltype
  method_plots
}
```

## 3. Get lineage-specific markers

Each branch needs its own early/terminal marker pair: a shared stem/
progenitor signature as the "early" set, and a lineage-specific terminal
signature per branch. `get_markers_msigdb()` pulls these by gene-set name,
and `filter_markers()` keeps only the genes well-detected in this dataset.

```{r markers}
ms_ery <- get_markers_msigdb(
  early    = "HAY_BONE_MARROW_CD34_POS_HSC",
  terminal = "WP_ERYTHROPOIESIS",
  species  = "Mus musculus"
)
ms_b <- get_markers_msigdb(
  early    = "HAY_BONE_MARROW_CD34_POS_HSC",
  terminal = "HADDAD_B_LYMPHOCYTE_PROGENITOR",
  species  = "Mus musculus"
)

fm_ery <- filter_markers(ms_ery, stem, top_n = 30, min_detection = 0.10)
fm_b   <- filter_markers(ms_b,   stem, top_n = 30, min_detection = 0.10)
```

## 4. Define the branch structure

Branched DOE functions take `early_markers_list`/`terminal_markers_list` as
**named lists keyed by branch name** — the union of these names defines the
branches BioTrajX will evaluate. `branch_filters` then tells BioTrajX which
`Phenotype` labels belong to each branch:

```{r branch-structure}
early_markers_list <- list(
  Stem_to_Ery = fm_ery$early,
  Stem_to_B   = fm_b$early
)
terminal_markers_list <- list(
  Stem_to_Ery = fm_ery$terminal,
  Stem_to_B   = fm_b$terminal
)

branch_filters <- list(
  Stem_to_Ery = list(
    include   = c("Stem_Progenitors", "Erythroid_progenitors_Erythroblasts", "Erythrocytes"),
    min_cells = 50
  ),
  Stem_to_B = list(
    include   = c("Stem_Progenitors", "Immature_B", "Mature_B"),
    min_cells = 50
  )
)
```

## 5. Evaluate a single pseudotime trajectory

```{r single-doe}
res_single <- compute_single_doe_branched(
  expr_or_seurat         = stem,
  pseudotime             = stem$Monocle3,
  early_markers_list     = early_markers_list,
  terminal_markers_list  = terminal_markers_list,
  cluster_labels         = "Phenotype",
  branch_filters         = branch_filters,
  E_method               = "gmm"
)

res_single$aggregate_DOE
plot(res_single, type = "bar")
plot(res_single, type = "radar")
plot(res_single, type = "heatmap")
```

## 6. Compare all eight methods, per branch

```{r multi-doe}
pt_list <- setNames(lapply(pt_cols, function(col) stem@meta.data[[col]]), pt_cols)
pt_list <- pt_list[sapply(pt_list, function(x) sum(!is.na(x)) > 10)]

res <- compute_multi_doe_branched(
  expr_or_seurat         = stem,
  pseudotime_list        = pt_list,
  early_markers_list     = early_markers_list,
  terminal_markers_list  = terminal_markers_list,
  cluster_labels         = "Phenotype",
  branch_filters         = branch_filters,
  E_method               = "gmm"
)

res$comparison_overall
plot(res, scope = "branch", type = "bar", branch_mode = "facet")
plot(res, scope = "branch", type = "radar", branch_mode = "facet")
plot(res, scope = "branch", type = "heatmap", branch_mode = "facet")
plot(res, scope = "overall", type = "bar")
plot(res, scope = "overall", type = "heatmap")
```

`scope = "overall"` only carries one score per trajectory (`aggregate_DOE`) — there's no per-metric D/O/E breakdown at that level to put on a radar's axes, so `type = "radar"` isn't available for `scope = "overall"`; use `scope = "branch"` for a per-metric radar instead.

## 7. Bonus: overall rank vs. per-branch rank

A method that looks best overall is not necessarily the best choice for
every branch — `compute_multi_doe_branched()` tracks branch-specific
rankings precisely so you can catch that:

```{r per-branch-rank}
res$comparison_overall
res$comparison_by_branch
res$best_per_branch
```

A method that wins on the erythroid branch can rank far lower on the B-cell
branch (and vice versa); the overall `aggregate_DOE` score is a cell-count-
weighted average across branches, so it can mask a method that's excellent
on one branch and poor on the other.
