---
title: "Validating BioTrajX Against Ground Truth (GSE131847 LCMV Time Course)"
output: github_document
vignette: >
  %\VignetteIndexEntry{Validating BioTrajX Against Ground Truth (GSE131847 LCMV Time Course)}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  warning = FALSE,
  message = FALSE,
  eval = FALSE   # requires downloaded data — run interactively after loading GSE131847_seu.rds
)
```

The linear-trajectory and branched-trajectory vignettes evaluate DOE against
literature marker genes and known cell-type clusters — proxies for "did this
method get the biology right," but not a direct check against ground truth
of *time*. This vignette uses a dataset where the real sampling time is
known: an LCMV infection time course (GSE131847), where CD8⁺ T cells were
collected at defined days post-infection (`naive`, `d3`, ..., `d90`). Since
the true ordering of cells is known here, we can ask directly: does a
method's DOE score actually track how well its pseudotime recovers the real
timeline?

As with the other two vignettes, pseudotime from eight TI methods is already
computed and stored as `meta.data` columns.

## 1. Load the data

```{r load}
library(BioTrajX)
library(Seurat)

gse <- readRDS("data/GSE131847_seu.rds")
DefaultAssay(gse) <- "SCT"
gse
```

`gse$cell_type` is not a cell *type* in the usual sense — it's the known
sampling day (`naive`, `d3`, `d4`, ..., `d90`), used here as ground truth
rather than as a biological cluster label. Setting the assay to `SCT`
ensures marker scoring and DOE both use SCTransform-normalized expression,
which is how this dataset was processed upstream.

## 2. Visualize sampling day and pseudotime

One UMAP panel for the known sampling day, plus one panel per TI method
colored by that method's pseudotime:

```{r visualize}
pt_cols <- c("Slingshot", "CytoTRACE", "Monocle3", "DPT",
             "SCORPIUS", "TSCAN", "PAGA-DPT", "Palantir")

gse$cell_type <- factor(
  gse$cell_type,
  levels = c("naive", "d3", "d4", "d5", "d6", "d7", "d10", "d14", "d21", "d32", "d60", "d90")
)

umap_df <- as.data.frame(Embeddings(gse, "umap"))
colnames(umap_df) <- c("UMAP1", "UMAP2")
umap_df$Day <- gse$cell_type[rownames(umap_df)]

p_day <- ggplot2::ggplot(umap_df, ggplot2::aes(UMAP1, UMAP2, colour = Day)) +
  ggplot2::geom_point(size = 0.5, alpha = 0.7) +
  ggplot2::theme_classic(base_size = 10) +
  ggplot2::labs(title = "Sampling day (ground truth)", colour = NULL)

method_plots <- lapply(pt_cols, function(method) {
  df <- cbind(umap_df[, c("UMAP1", "UMAP2")], Pseudotime = gse@meta.data[[method]])
  ggplot2::ggplot(df, ggplot2::aes(UMAP1, UMAP2, colour = Pseudotime)) +
    ggplot2::geom_point(size = 0.5, alpha = 0.7) +
    ggplot2::scale_colour_distiller(palette = "Spectral", direction = -1,
                                     na.value = "grey80") +
    ggplot2::theme_classic(base_size = 10) +
    ggplot2::labs(title = method)
})

if (requireNamespace("patchwork", quietly = TRUE)) {
  patchwork::wrap_plots(c(list(p_day), method_plots), ncol = 3)
} else {
  p_day
  method_plots
}
```

## 3. Markers and multi-method DOE

```{r markers-doe}
marker_set <- get_markers_msigdb(
  early    = "GSE41867_NAIVE_VS_DAY8_LCMV_EFFECTOR_CD8_TCELL_UP",
  terminal = "GSE41867_NAIVE_VS_DAY8_LCMV_EFFECTOR_CD8_TCELL_DN",
  species  = "Mus musculus"
)
filtered <- filter_markers(marker_set, gse, top_n = 30, min_detection = 0.10)

pt_list <- setNames(lapply(pt_cols, function(col) gse@meta.data[[col]]), pt_cols)
pt_list <- pt_list[sapply(pt_list, function(x) sum(!is.na(x)) > 10)]

res <- compute_multi_doe_linear(
  expr_or_seurat   = gse,
  pseudotime_list  = pt_list,
  early_markers    = filtered$early,
  terminal_markers = filtered$terminal,
  cluster_labels   = "cell_type",
  E_method         = "gmm"
)
```

This dataset has ~26,000 cells, so `compute_multi_doe_linear()` can take a
few minutes and several GB of free memory to run across all eight methods —
that's expected, not an error.

## 4. Visualize and rank

```{r visualize-rank}
res$comparison_summary
res$best_trajectory
plot(res, type = "bar")
plot(res, type = "radar")
plot(res, type = "heatmap")
```

## 5. Bonus: does DOE_score track ground-truth recovery?

DOE never sees the sampling day — it only sees marker genes, pseudotime, and
(optionally) cluster labels. If DOE is doing its job, methods it scores
highly should also be the ones whose pseudotime best recovers the real day
ordering:

```{r ground-truth-check}
day_lookup <- c(naive = 0, d3 = 3, d4 = 4, d5 = 5, d6 = 6, d7 = 7,
                 d10 = 10, d14 = 14, d21 = 21, d32 = 32, d60 = 60, d90 = 90)
true_day <- day_lookup[as.character(gse$cell_type)]

recovery <- sapply(pt_list, function(pt) {
  suppressWarnings(cor(pt, true_day, method = "spearman", use = "complete.obs"))
})

comparison <- res$comparison_summary
comparison$day_recovery <- recovery[comparison$trajectory]
comparison[, c("trajectory", "DOE_score", "day_recovery")]

cor.test(comparison$DOE_score, comparison$day_recovery, method = "pearson")
```

`day_recovery` is kept signed (not absolute) deliberately: a method run
backwards gets a negative Spearman rho against true day, and a trajectory
with a genuinely wrong orientation should score low on DOE too — taking the
absolute value would hide exactly the failure mode DOE is meant to catch.
The strong positive correlation here (DOE score tracks day-recovery) is the
real validation of DOE as a label-free proxy for ground truth.
